
```{r}
library(tidyverse)
library(vegan)
set.seed(12345) 
library(ggthemes)
```


Mantel and NMDS of rarefied table. 

```{r}
asv <- read_csv("1.3_ITS_ASV_Table.csv")
metadata <- read_csv("1.1_Metadata.csv") 
```
I was not able to rarefy my ITS table as I would have lost too many samples, therefore, I chose to use a square root transformation to normalize my table instead. 
```{r}
# Apply square root transformation to all sample columns, keeping ASV intact
asv <- asv %>%
  mutate(across(-c(ASV, Taxonomy), sqrt))
```

Bray-curtis cannot handle negative values, scale variables with negatives.
Chose to use Min-Max Normalization to make all values non-negative while preserving relative differences.
```{r}
# Clean and scale metadata
metadata <- metadata %>%
  select(Sample, depth, depth_num, Depth_Practice, Location, Adaptive_Score, Practice, 
         Location_Depth, Biomass, percent_clay, percent_sand, percent_silt, pH, 
         total_c, maom_c, pom_c, total_n, maom_n, pom_n, c_n_ratio, mean_annual_temp, 
         mean_annual_precip, radiation, PET_mm_year) %>%
  mutate(mean_annual_temp_scaled = (mean_annual_temp - min(mean_annual_temp, na.rm = TRUE)) / 
                                   (max(mean_annual_temp, na.rm = TRUE) - min(mean_annual_temp, na.rm = TRUE))) %>%
  select(-mean_annual_temp) %>%
  drop_na()
```


We need to invert the dataframe so that each row is a sample and each column is an ASV. This is also an opprotunity to filter out any samples/treatments we are not interested in.
```{r}
flip <- asv %>%
  select(-Taxonomy) %>%
  pivot_longer(-ASV, names_to = "Sample", values_to = "count") %>%
  pivot_wider(names_from = ASV, values_from = count, values_fill = 0) %>%  # Fill missing counts with 0
  inner_join(metadata, by = "Sample") %>%  # Ensure you join by 'Sample'
  filter(!Sample %in% c(
    "WYCO_0923_05_PR_D3_B_C03_XT",
    "WYCO_0923_05_PR_D2_B_C03_XT",
    "WYCO_0923_015_AD_D3_B_C05_XT"
  ))
```

Make sure that the dataframe and metadata have the same number of rows and that the rows are in the exact same order.
```{r}
meta <- flip %>% 
  select(1) %>% 
  inner_join(metadata)
```

Here we filter the dataframe to remove the first column composed of sample names.
```{r}
table <- flip %>% 
 select(2:3266) 
```


```{r}
# Ensure only numeric columns are in the table before calculating Bray-Curtis
table_numeric <- table %>%
  select(where(is.numeric)) %>%  # Select only numeric columns
  drop_na()
```



The next step is to take the data frame and turn it into a distance matrix. Here we are calculating distance using Bray-Curtis distance, but many other methods are available. Distance matrices are pairwise comparisons that compare the similarity of every community on a scale from 0 to 1. 0 indicates communities are exactly the same and 1 indicates there is no similarity in the communities.
```{r}
data.t <- as.matrix(vegdist(table_numeric, method = 'bray'))
```

Running a mantel test on NMDS
```{r}
# Step 6: Run Mantel Tests
# Function for running mantel tests
run_mantel_test <- function(variable, distance_matrix) {
  variable_dist <- dist(variable, method = "euclidean")
  mantel_test <- mantel(distance_matrix, variable_dist, method = "spearman", permutations = 999)
  return(list(statistic = mantel_test$statistic, p_value = mantel_test$signif))
}

# Mantel tests for environmental variables
mantel_results <- list(
  "pH" = run_mantel_test(meta$pH, data.t),
  "SOC" = run_mantel_test(meta$total_c, data.t),
  "MAOM C" = run_mantel_test(meta$maom_c, data.t),
  "POM C" = run_mantel_test(meta$pom_c, data.t),
  "Nitrogen" = run_mantel_test(meta$total_n, data.t),
  "MAOM N" = run_mantel_test(meta$maom_n, data.t),
  "POM N" = run_mantel_test(meta$pom_n, data.t),
  "Clay" = run_mantel_test(meta$percent_clay, data.t),
  "Sand" = run_mantel_test(meta$percent_sand, data.t),
  "Silt" = run_mantel_test(meta$percent_silt, data.t),
  "Depth" = run_mantel_test(meta$depth_num, data.t),
  "Adaptive Score" = run_mantel_test(meta$Adaptive_Score, data.t),
  "MAT" = run_mantel_test(meta$mean_annual_temp_scaled, data.t), 
  "MAP" = run_mantel_test(meta$mean_annual_precip, data.t), 
  "PET" = run_mantel_test(meta$PET_mm_year, data.t), 
  "Solar Radiation" = run_mantel_test(meta$radiation, data.t)
)

# Create a dataframe of Mantel results
mantel_df <- data.frame(
  variable = names(mantel_results),
  rho_value = sapply(mantel_results, function(x) x$statistic),
  p_val = sapply(mantel_results, function(x) x$p_value)
)
```


```{r}
# Save Mantel test results
#write.csv(mantel_df, file = "/Users/lmoore/Desktop/Documents/CSU/PhD/Dissertation/Grazing Microbiome/Data_Figures/Mantel/ITS_mantel_results_rarefied.csv")
```

Create bar chart of mantel test
```{r}
# Ensure the variables are ordered by rho_value in descending order for better visualization
mantel_df <- mantel_df %>%
  mutate(variable = reorder(variable, -rho_value),
         significance = case_when(
           p_val <= 0.001 ~ "***",
           p_val <= 0.01  ~ "**",
           p_val <= 0.05  ~ "*",
           TRUE ~ ""
         ))  # Create a column for significance stars

# Create the bar chart
mantel_plot <- ggplot(mantel_df, aes(x = variable, y = rho_value)) +
  geom_bar(stat = "identity", fill = "#896A78") +  # Create bar chart
  geom_text(aes(label = significance), vjust = -0.5, size = 4) +  # Display significance stars above bars
  labs(title = "Mantel Test Results", x = "Environmental Variables", y = "Mantel's Rho") +  # Axis and title labels
  theme_tufte() +  # Clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))   # Rotate x-axis labels

print(mantel_plot)

```

We can do pairwise comparisons to understand which sites are different from each other
```{r}
#make a new table of all of these with just the individual sites and their depths
library(devtools)
#install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis)
pairwise.adonis<-pairwise.adonis2(data.t ~ Location_Depth, data=meta, permutations = 999)
pairwise.adonis
```

Now we can visualize community similarity using NMDS (non-metric multidimensional scaling), an iterative and non-parametirc method which aims to plot pairwise dissimilarity between communities. Each point will represent once sample (or microbial community). The closer points are together the more similar those communities are to each other.

How good an NMDS visualization is at representing your community can be assessed by the stress values. Typically NMDS with stress below 0.2 can be retained.
Stress value below 0.05 is considered a great fit
Stress value below 0.1 is considered a good fit 
Stress value below 0.2 is considered an ok fit 
A stress value above 0.2 is questionable and anything approaching 0.3 is considered arbitrary

Importatant parameters in the metaMDS function:
"trymax"" is the maximum number of random starts it will perform to try to reach convergence. The defaul value is 20 and if it reaches convergence after 20 random starts it will end there. Otherwise it will go either until it reaches convergence or the number of tries youhave put in.

"k" is the number of dimensions. If you are having trouble converging then you can try increasing the k value. If you don't set k it will run with 2 dimensions. Increasing hte k value will likely increase the stress. Anything above 5 dimensions makes interpreting 2-dimensional plots difficult.



```{r}
set.seed(12345)
NMDS<-metaMDS(data.t, trymax= 200)
MDS1 = NMDS$points[,1]
MDS2 = NMDS$points[,2]
NMDS = data.frame(NMDS1 = MDS1, NMDS2 = MDS2)


```


```{r}
mrpp.1 <- mrpp(data.t, meta$Practice, permutations=999, distance="bray")
mrpp.1
```


```{r}
anosim.1 <- anosim(data.t, meta$Practice, permutations=999, distance="bray")
anosim.1
```



```{r}
meta$Depth_Practice <- factor(meta$Depth_Practice, levels = c('D1_AMP','D2_AMP', 'D3_AMP', 'D1_Conventional', 'D2_Conventional', 'D3_Conventional' ))
NMDS_plot <- ggplot(NMDS) +
  geom_point(aes(x=NMDS1, y=NMDS2, col=meta$Depth_Practice,  shape=meta$Location), size=4) + 
  theme_bw() +
  labs(color = "Depth and Practice", shape= "Site")+
   scale_shape_manual(values = c(15,16,17,18,4,7)) +  #change shape of points
  #stat_ellipse(aes(x=NMDS1, y=NMDS2, col=meta$Depth_Practice)) + #add ellipse around variables
  scale_color_manual(values=c('#b35806','#f1a340','#fee0b6', '#542788','#998ec3', '#d8daeb'))+ #change color of points
   theme(text=element_text(size=21))
 NMDS_plot

```








