

```{r}
library(tidyverse)
library(vegan)
library(plyr)
library(car)
```


Read in data
```{r}
asv <- read.csv("1.2_16S_ASV_Table.csv")
metadata <- read.csv("1.1_Metadata.csv")

```

Clean up data
```{r}
# Remove the 'Taxon' column from the 'asv' data frame and create 'asv_new'
asv_new <- asv %>% 
  select(-Taxon)
# Remove the columns from the 'metadata' dataframe
metadata <- metadata %>% 
  select(Sample, maom_c, Practice, depth) %>% 
  drop_na()

```


```{r}
table <- asv_new %>%
  pivot_longer(cols = -ASV,  # This selects all columns except the ASV column
               names_to = "Sample", 
               values_to = "counts") %>% 
  pivot_wider(names_from = ASV, values_from = counts)


```

Apply the richness function
```{r}
richness <- ddply(table, ~Sample, function(x) {
  data.frame(Richness = sum(x[-1] > 0))
})

```

Apply the evenness function
```{r}
evenness <- ddply(table, ~Sample, function(x) {
  data.frame(Evenness = diversity(x[-1], index = "simpson") / log(sum(x[-1] > 0)))
})

```


Apply the shannon's diversity function
```{r}
shannon <- ddply(table, ~Sample, function(x) {
  data.frame(Shannon = diversity(x[-1], index = "shannon"))
})

```

Join data
```{r}
alpha_diversity <- shannon %>% 
  inner_join(richness, by = "Sample") %>% 
  inner_join(evenness, by = "Sample") %>% 
  inner_join(metadata, by = "Sample")

```




Plot the relationship between management category (practice) and Alpha diversity
```{r}
#plot for practice vs all 3 depths
plot2 <- alpha_diversity %>%
  pivot_longer(cols = c(Richness, Evenness, Shannon), names_to = "Diversity_Type", values_to = "Diversity") %>%
  ggplot(aes(x = depth, y = Diversity, fill = Practice)) +
  geom_boxplot() +
  facet_wrap(~Diversity_Type, scales = "free") +
  theme_bw()

plot2

```



Statistical testing for Alpha Diversity and Management Category (Practice)
```{r}
# Linear regression for each diversity metric
lm_richness2 <- lm(Richness ~ Practice, data = alpha_diversity)
lm_evenness2 <- lm(Evenness ~ Practice, data = alpha_diversity)
lm_shannon2 <- lm(Shannon ~ Practice, data = alpha_diversity)

# View the summary of the regression to see p-values and R-squared values
summary(lm_richness2)  # For Richness
summary(lm_evenness2)  # For Evenness
summary(lm_shannon2)   # For Shannon

# ANOVA to test if Shannon diversity differs significantly by Practice
res_aov <- aov(Shannon ~ Practice, data = alpha_diversity)
summary(res_aov)

# Shapiro-Wilk test for normality of residuals from ANOVA (required for ANOVA validity)
shapiro.test(res_aov$residuals)

# Kruskal-Wallis test as a non-parametric alternative to ANOVA (for non-normal residuals)
kw_test <- kruskal.test(Shannon ~ Practice, data = alpha_diversity)
kw_test

# Wilcoxon rank sum test for pairwise differences (used when ANOVA assumptions are violated)
pairwise.wilcox.test(alpha_diversity$Shannon, alpha_diversity$Practice,
                 p.adjust.method = "BH")


```


Plot the relationship between MAOC and Alpha Diversity
```{r}
plot_maoc_vs_diversity <- alpha_diversity %>%
  pivot_longer(cols = c(Richness, Evenness, Shannon), names_to = "Diversity_Type", values_to = "Diversity") %>%
  ggplot(aes(x = maom_c, y = Diversity)) +
  geom_point(aes(color = Practice, shape = Practice)) +    # Color and shape by Practice
  geom_smooth(method = "lm", se = FALSE, color = "gray30") +  # Black regression line
  facet_wrap(~Diversity_Type, scales = "free") +
  scale_color_manual(values = c("Adaptive" = "#be7245", "Conventional" = "#998ec3")) +
  theme_bw() +
  labs(
    x = "Mineral-Associated Organic Carbon (mg C/g Soil)",
    y = "Alpha Diversity",
    title = "Relationship between Alpha Diversity and MAOC"
  )

plot_maoc_vs_diversity


```

Statistical testing for MAOC and Alpha Diversity
```{r}
# Linear regression for each diversity metric
lm_richness <- lm(Richness ~ maom_c, data = alpha_diversity)
lm_evenness <- lm(Evenness ~ maom_c, data = alpha_diversity)
lm_shannon <- lm(Shannon ~ maom_c, data = alpha_diversity)

# View the summary of the regression to see p-values and R-squared values
summary(lm_richness)  # For Richness
summary(lm_evenness)  # For Evenness
summary(lm_shannon)   # For Shannon

# Shapiro Wilks tests for normality of residuals from linear models (whether residuals from lm models are normally distributed)
shapiro.test(residuals(lm_shannon))
shapiro.test(residuals(lm_evenness))
shapiro.test(residuals(lm_richness))

# Spearman correaltion - used when residuals from linear regression are not normally distributed
cor.test(alpha_diversity$Shannon, alpha_diversity$maom_c, method = "spearman")
cor.test(alpha_diversity$Evenness, alpha_diversity$maom_c, method = "spearman")


```
