

```{r}
library(plyr)
library(tidyverse)
library(vegan)
library(car)

```
Load Data
```{r}
asv <- read_csv("1.3_ITS_ASV_Table.csv")
metadata <- read.csv("metadata.csv")
```
Clean Data
```{r}
#Remove taxon column
asv <- asv %>% 
  select(-Taxon)

# Choose metadata variables
metadata <- metadata %>% 
  select(Sample, maom_c, Practice, depth) %>% 
  drop_na()
```


```{r}
table <- asv %>%
  pivot_longer(cols = -ASV,  # This selects all columns except the ASV column
               names_to = "Sample", 
               values_to = "counts") %>% 
  pivot_wider(names_from = ASV, values_from = counts)


```

Square Root Transform the ASV Table
```{r}
# Apply square root transformation to all ASV columns, keeping Sample column unchanged
sqrt_table <- table %>%
  mutate(across(-Sample, sqrt))


# Apply relative abundance normalization only to ASV columns, keeping Sample column unchanged
relab_table <- sqrt_table %>%
  mutate(across(-Sample, ~ . / sum(.), .names = NULL))
```

Apply richness function
```{r}
richness <- ddply(sqrt_table, ~Sample, function(x) {
  data.frame(Richness = sum(x[-1] > 0))
})

```

Apply evenness function
```{r}
evenness <- ddply(sqrt_table, ~Sample, function(x) {
  data.frame(Evenness = diversity(x[-1], index = "simpson") / log(sum(x[-1] > 0)))
})

```
Apply Shannon's Diveristy Function
```{r}
shannon <- ddply(sqrt_table, ~Sample, function(x) {
  data.frame(Shannon = diversity(x[-1], index = "shannon"))
})

```

Join Data
```{r}
alpha_diversity <- shannon %>% 
  inner_join(richness, by = "Sample") %>% 
  inner_join(evenness, by = "Sample") %>% 
  inner_join(metadata, by = "Sample")
```




Plot the relationship between management category (practice) and Alpha diversity
```{r}
#plot for practice vs all 3 depths
plot2 <- alpha_diversity %>%
  pivot_longer(cols = c(Richness, Evenness, Shannon), names_to = "Diversity_Type", values_to = "Diversity") %>%
  ggplot(aes(x = depth, y = Diversity, fill = Practice)) +
  geom_boxplot() +
  facet_wrap(~Diversity_Type, scales = "free") +
  theme_bw()

plot2

```



Statistical testing for Alpha Diversity and Management Category (Practice)
```{r}
# Linear regression for each diversity metric
lm_richness <- lm(Richness ~ Practice, data = alpha_diversity)
lm_evenness <- lm(Evenness ~ Practice, data = alpha_diversity)
lm_shannon <- lm(Shannon ~ Practice, data = alpha_diversity)

# View the summary of the regression to see p-values and R-squared values
summary(lm_richness)  # For Richness
summary(lm_evenness)  # For Evenness
summary(lm_shannon)   # For Shannon

# Shapiro-Wilk test for normality of residuals
shapiro.test(residuals(lm_shannon))

# Levene's test for homogeneity of variance
leveneTest(Shannon ~ Practice, data = alpha_diversity)

# Kruskal-Wallis test for if residuals are non-normal or if you want to confirm results
kruskal.test(Shannon ~ Practice, data = alpha_diversity)
```


Plot the relationship between MAOC and Alpha Diversity
```{r}
# Plot alpha diversity vs maaom_c (MAOC) for all diversity types
plot_maoc_vs_diversity <- alpha_diversity %>%
  pivot_longer(cols = c(Richness, Evenness, Shannon), names_to = "Diversity_Type", values_to = "Diversity") %>%
  ggplot(aes(x = maom_c, y = Diversity, color = Diversity_Type)) +
  geom_point(aes(shape = Practice)) +     # Scatter plot with points colored by Diversity_Type
  geom_smooth(method = "lm", se = FALSE) + # Add a linear regression line
  facet_wrap(~Diversity_Type, scales = "free") +  # Separate plots for each diversity metric
  theme_bw() +
  labs(x = "Mineral-Associated Organic Carbon (mg C/ g soil)", y = "Alpha Diversity", title = "Relationship between Alpha Diversity and MAOC")

plot_maoc_vs_diversity

```



Statistical testing for MAOC and Alpha Diversity
```{r}
# Linear regression for each diversity metric
lm_richness2 <- lm(Richness ~ maom_c, data = alpha_diversity)
lm_evenness2 <- lm(Evenness ~ maom_c, data = alpha_diversity)
lm_shannon2 <- lm(Shannon ~ maom_c, data = alpha_diversity)

# View the summary of the regression to see p-values and R-squared values
summary(lm_richness2)  # For Richness
summary(lm_evenness2)  # For Evenness
summary(lm_shannon2)   # For Shannon


# Shapiro-Wilk normality test on residuals from each model
shapiro.test(residuals(lm_richness2))   # For Richness ~ maom_c
shapiro.test(residuals(lm_evenness2))   # For Evenness ~ maom_c
shapiro.test(residuals(lm_shannon2))    # For Shannon ~ maom_c

# Spearman's rank correlation (non-parametric alternative) for residuals not normally distributed
cor.test(alpha_diversity$Evenness, alpha_diversity$maom_c, method = "spearman")
cor.test(alpha_diversity$Shannon, alpha_diversity$maom_c, method = "spearman")
```
